---
title: 'Case Study: Modelling the progression of migration in knee prostheses'
author: 'Yuxuan LI (s2485265), Ariel Liang (s2614693), Yuying Tan (s2586401), María Zilli (s2261030)'
date: "May 1st, 2020"
output:
  html_document
geometry: margin=0.8in
fontsize: 12pt
---

# Introduction 

The following case study intends to analyse the data set from a longitudinal study on patients who had a knee operation, where a new knee protheses was tested. The outcome variable of the analysis is the Maximum Total Point Motion (MTPM) measured in mm. 
The time frame from this data is between March 2009 and July 2010 from 60 patients, and 5 measurements were recorded: the first day after the surgery (taken as baseline for this analysis), then at 3 months, 1 year, 2 years and at 5 years postop- eratively. In addition to MTPM, the age, gender and the BMI of the patients is also recorded.

A multivariate model and a linear mixed effects model are considered. 


# Data preparation 

```{r}
# define a function that look beforehead if come across NA values
pre_val <- function(vector, location)
{
  if (is.na(vector[location]))
  {
    return(pre_val(vector, location-1))
  } else
  {
    return(vector[location])
  }
}


# define a function that look afterhead if come across NA values
post_val <- function(vector, location)
{
  if (!is.na(vector[location]))
  {
    return(vector[location])
  } else if (location < length(vector))
  {
    return(post_val(vector, location+1))
  } else
  {
    return(pre_val(vector, location-1))
  }
}

#load data
#file. <- "..." # specify the directory with data
DATA <- read.spss(file.path(file., "Prostheses-2020.txt"), to.data.frame = TRUE)

#delete NA values
n_patient <- max(DATA$ID)
# substitute NA values with recent obs of the same patient
for (id in 1:n_patient)
{
  # extract patient data one by one
  data <- DATA[DATA$ID==id,]
  # operate while there're NA values
  if (sum(as.numeric(is.na(data$nMTPM)))>0)
  {
    row_na_list <- which(is.na(data$nMTPM))
    for (row_na in row_na_list)
    {
      if (row_na==1)
      {
        # if the 1st obs is missing, substitute with subsequent values
        data$nMTPM[row_na] <- post_val(data$nMTPM, row_na + 1)
      } 
      else if (row_na==dim(data)[1])
      {
        # if the last obs is missing, substitute with previous values
        data$nMTPM[row_na] <- pre_val(data$nMTPM, row_na - 1)
      }
      else 
      {
        # in other cases, substitue NA with the average of recent obs
        a <- pre_val(data$nMTPM, row_na - 1)
        b <- post_val(data$nMTPM, row_na + 1)
        data$nMTPM[row_na] <- (a+b)/2
      }
    }
  }
  DATA <- rbind(DATA,data)
}
# check if there're any more NA values
k_row <- dim(DATA)[1]
data <- DATA[(k_row/2+1):k_row,]
if (sum(as.numeric(is.na(data)))==0)
{
  DATA <- data
  save(DATA, file="prostheses.RData")
  knitr::kable(DATA[1:10,], row.names=FALSE, align="lrcrcrrr", digits=4,
               caption="Data of the First Two Patients")
} else
{
  print("clean the data again.")
}
# remove redundant variables
rm(data, id, k_row, row_na, row_na_list, a, b)
```

**The workspace made be loaded from prostheses.RData in case the codes failed**   
load("prostheses.RData") 

**The following libraries were used**

require(pacman)
library(lattice)
library(lme4)
library(nlme)
library(ggplot2)

**The following variables were transformed for the anlysis:** 

+ Transformation of the variable type of protheses to factor levels 1 and 2 
+ Transformation of the variable Months into factor, the following analysis uses the variable Months instead of Years as its interpretation is easier. The baseline is considered to be month 0 equal to the first MTPM one day after the operation. 

```{r}
DATA$Month.f <- factor(DATA$FU.Months, levels = c(0,3,12,24,60))
DATA$Type.f <- factor(DATA$Type, levels = c(1,2))
```

# Q. 1 

## Plots for Mean Structure and Data exploration 

The following section intends to describe the data, and use graphical techniques to explore the mean structure, the variance structure and the correlation structure. 

**Mean per protheses type and month** 
```{r}
with(DATA, tapply(nMTPM, list(Month.f, Type), mean, na.rm = TRUE))
```
The means for the Maximum Total Point Motion (MTPM) both increase with both types of prothesis, but the change in mean is larger for type 2. 

**Variance per protheses type and month** 
```{r}
with(DATA, tapply(nMTPM, list(Month.f, Type), var, na.rm = TRUE))
```
Variance increases for both types with time, however there are larger variance values for type 2. 

**Plot for individual profiles** 

```{r}
xyplot(nMTPM ~ Month.f , groups =ID, data = DATA, xlab = "Month", ylab = "MTPM(mm) ", type = "l")
```

Plot of individual profiles shows an increase in MTPM for almost all individuals from baseline (first day after the cirgury) to 3 months after the operation, after the 3 months we see a fairly continuos trend. 
However, some outliers appear in the graph, particularly individuals which show a great increase in mobility throughout the months and others which show a decerase in mobility. 

**Plot showing individual profiles per type of prothesis** 
```{r}
xyplot(nMTPM ~ Month.f | Type.f, groups =ID, data = DATA, xlab = "Month", ylab = "MTPM (mm)", type = "l")
```

For the type 2 we see a steep increase in MTPM 3 months after the cirgury. After the three months more steady however differences in individual profiles.For type 1 there is a slower increase in the profiles and more diverse MTPM. 

**Plot for mean MTPM for each type and period, for both types the MTPM increase**
```{r}
xyplot(nMTPM ~ Month.f | Type.f, groups =ID, data = DATA, xlab = "Month", ylab = "MTPM (mm)", type = "l", panel = function(x, y) {
  panel.average(x, y, horizontal = FALSE, col = "blue")
})
```

The mean shows a steap increase for type 2 protheses for the first 3 months postoperation and after a flatter increase, while for type 1 the increase in MTPM is more constant. Further, type 1 shows a lower MTPM mean level than type 2. 


**Age vs. MTPM**
```{r}
p <- ggplot(data = DATA, aes(x = Age, y = nMTPM)) + geom_point( col = "darkblue", alpha = .15)
p+geom_smooth(method = lm )
```

The graph shows slightly higher values of MTPM for age below 50, however there are only five observations from patients below 50 years old. The patients are more clustered between 60 and 80 years (this is expected from the nature of the study). From 50 to 80 years there is no clear relationship between age and MTPM levels. 


**Body Mas Index vs MTPM** 

```{r,echo=FALSE}
p1 <- ggplot(data = DATA, aes(x = BMI, y = nMTPM)) + geom_point( col = "darkblue", alpha = .15)
p1 + geom_smooth(method = lm )
```


There is no clear relationship between BMI and MTPM, even though the graph shows a slight increase, there is a cluster of patients between the levels of 25 to 30 BMI. 

**Sex vs.MTPM** 
```{r}
plot(nMTPM ~ Sex , data = DATA, type = "l", groups = ID, col = c("darkblue", "darkred"))
```


MTPM values seem to have more variation for females than to males. The median is lower for males than females, as well as the upper values (larger values for MTPM correspond to females). 



**Plot per protheses type**


```{r,echo=FALSE}
p2 <- ggplot(DATA, aes( x = Month.f, y=nMTPM , group= ID )) 
p2 + geom_point(aes(colour= Type.f)) 
```


More values are clustered for type 1 at lower values of MTPM and higher values for type 2. 


**Plot per Sex and Type**


```{r,echo=FALSE}
p2 + geom_point(aes(colour= Type.f)) + facet_grid(.~Sex) 
```


For males type 1 values look clustered around lower values of MTPM, for females there is more variation. Further, females show higher values of MTPM than males.  


**Interaction between Sex and Age**


```{r}
p3 <- ggplot(data = DATA, aes(x = Age, y = nMTPM,fill = Sex)) + geom_point( col = "darkblue", alpha = .15)
p3+geom_smooth(method = lm )
``` 

**Interaction between BMI and Age**

```{r}
p4 <- ggplot(data = DATA, aes(x = BMI, y = nMTPM,fill = Sex)) + geom_point( col = "darkblue", alpha = .15)
p4+geom_smooth(method = lm )
``` 

**Interaction between Sex and nMTPM**
```{r}
p5 <- ggplot(data = DATA, aes(x = FU.Months, y = nMTPM,fill = Sex)) + geom_point( col = "darkblue", alpha = .15)
p5+geom_smooth(method = lm )
```

##Conclusions
From the spaghetti plots we can assume that there is a linear relationship between time and MTPM output variable. The exploratory analysis shows there are differences between the two types of protheses, and different migrations for men and women, however this difference might be from the data itself. 

# Q.2 Multivariate model 

**1.First we consider an elaborate model for the mean response** 

From the EDA, we consider the structure below for now. We will choose between an Unstructured model and (Semi-) Parametric model. 

```{r, eval=FALSE}
model1 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              method = "REML")


model2 <- gls(nMTPM ~ Age*Sex+FU.Month*Type+Sex*BMI+Sex*FU.Month,
              data = DATA,
              method = "REML")


summary(model1)
summary(model2)
model1 <- update(model1, method = "ML")
model2 <- update(model2, method = "ML")
anova(model1,model2)
```

Model1 is better than model2, we prefer Unstructured model over (Semi-) Parametric model 

**2.Select a correlation structure (same mean model but different covariance structures)**  

The following models consider compound symmetry, continuos ARI, linear Gaussian and constant variance in time strucutres. 

```{r, eval=FALSE}
model3 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corSymm(form = ~ 1 | ID),
              weights = varIdent(form = ~ 1 | FU.Months),
              na.action = na.exclude, method = "REML")
summary(model3)


cov.mat <- getVarCov(model3, individual = 3)
cov.mat
cov2cor(cov.mat) #We observe the  correlations decrease with time, so we try other correlation structures.

model4 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corCompSymm(form = ~ FU.Months | ID),
              na.action = na.exclude, method = "REML")

model5 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corAR1(form = ~ FU.Months| ID),
              na.action = na.exclude, method = "REML")

model6 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corLin(form = ~ FU.Months | ID),
              na.action = na.exclude, method = "REML")

model7 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corGaus(form = ~ FU.Months | ID),
              na.action = na.exclude, method = "REML")

model8 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corExp(form = ~ FU.Months | ID),
              na.action = na.exclude, method = "REML")

#Compare models 
anova(model4,model5, model6, model7, model8) 
#Model4 gives lowest c AIC and BIC (model with Compound Symmetry Structure)
```

**Extension of the correlation structures by assumption of heteroscedastic errors** 

```{r, eval= FALSE}
model9 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corCompSymm(form = ~ FU.Months | ID),
              weights =  varExp(form = ~ FU.Months),
              na.action = na.exclude, method = "REML")


model10 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corAR1(form = ~ FU.Months | ID),
              weights =  varExp(form = ~ FU.Months),
              na.action = na.exclude, method = "REML")


model11 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
              data = DATA,
              correlation = corLin(form = ~ FU.Months | ID),
              weights =  varExp(form = ~ FU.Months),
              na.action = na.exclude, method = "REML")


model12 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
               data = DATA,
               correlation = corGaus(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")


model13 <- gls(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f,
               data = DATA,
               correlation = corExp(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")

#Compare Models with heteroscedastic errors

anova(model9, model10, model11, model12, model13)

## BIC and AIC lowest for model 9 (Compound Symmetry Structure)

## We can compare model 4 and model 9 
anova(model9, model4)
## model 9 gives the lowest AIC & BIC so we proceed with model 9
```

**3.Now let us simplify the mean model by excluding non-significant terms** 

```{r, eval=FALSE}
summary(model9)

#exclude term Age
model14 <- update(model9, .~.-Age)
summary(model14)
anova(model14)

#exclude interaction between Sex and Age as the p-value shows the interaction is not significant (.844)

model15 <- gls(nMTPM ~ Month.f*Type.f+Sex*BMI+Sex*Month.f,
               data = DATA,
               correlation = corCompSymm(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")

summary(model15)
anova(model15)

#remove interaction between Sex and BMI
model16 <- gls(nMTPM ~ Month.f*Type.f+Sex*Month.f,
               data = DATA,
               correlation = corCompSymm(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")

summary(model16)
anova(model16)
```

```{r}
#remove Sex as results show it is not significant 
model17 <- gls(nMTPM ~ Month.f*Type.f+Sex*Month.f-Sex,
               data = DATA,
               correlation = corCompSymm(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")


summary(model17)
anova(model17)
```

##Residual plots

**Standarized residuals**
```{r}
plot(model17, resid(., type = "p") ~ fitted(.),type = c("p", "smooth"), lwd = 3)
```

**Standarized residuals per type** 
```{r}
plot(model17, resid(., type="p") ~ fitted(.)|Type.f,type = c("p", "smooth"), lwd = 3)
```

**Normalized residuals** 
```{r}
plot(model17, resid(., type="n") ~ fitted(.),type = c("p", "smooth"), lwd = 3)
```

**Normalized residuals**
```{r}
plot(model17, resid(., type="n") ~ fitted(.)|Type.f,type = c("p", "smooth"), lwd = 3)
```

#the variance is not constant


**a) Do both types migrate similarly over time?**
+ The values taken at baseline are not significant since the MTMP was measured 1 day after the operation,therefore there is no change of MTMP. 
+ The protheses types do not migrate similarly over time. Even though the general trend is an increase in MTPM for the first three months and after a stabilization, there are differences. For type 1, the mean for MTMP shows an increase from baseline until 5 years, and for type 2 there is an increase from baseline to the 3 months measurement and after a stabilization. The increase from baseline to 3 months is more significant for type 2. 

**b) Report the differences in the mean migration between the two operation types at each time point**
**Intervals for 95% confidence level**

```{r}
intervals(model17)
```
For type 1 the larger differences are between baseline and 3 months and from 3 months to 1 year after the operation, after this there is a mean increase of MTMP but it is more steady. 

For type 2 the larger difference in mean MTMP is between baseline and the first 3 months, after there is a slight decrease in MTMP. 

**c) Report the size of the migration change from 3 months to 1 year, 3 months to 2 years, 3 months to 5 years and change from 1 year to 5 year pero operation type and make statistical inference. Given the corresponding 95% confidence interval and p-value.**

For this we estimate the difference for each period separately with a t.test for each period 

```{r}
t.test(nMTPM ~ Type.f, data = DATA, subset = which (DATA$FU.Months== 0), paired =FALSE)
t.test(nMTPM ~ Type.f, data = DATA, subset = which (DATA$FU.Months== 3), paired =FALSE)
t.test(nMTPM ~ Type.f, data = DATA, subset = which (DATA$FU.Months== 12), paired =FALSE)
t.test(nMTPM ~ Type.f, data = DATA, subset = which (DATA$FU.Months== 24), paired =FALSE)
t.test(nMTPM ~ Type.f, data = DATA, subset = which (DATA$FU.Months== 60), paired =FALSE)
```

The p-values for each t test are significant except for the baseline which is one day after the operation. The type of protheses is statistically significant at all other periods separately. 

```{r}
#Type 1
data1 <- filter(DATA, Type.f == "1")
D1 <- data1[data1$Month.f == 3, "nMTPM"] - data1[data1$Month.f == 60, "nMTPM"]

#Type 2
data2 <- filter(DATA, Type.f == "2")
D2 <- data2[data2$Month.f == 3, "nMTPM"] - data2[data2$Month.f == 60, "nMTPM"]

 t.test(x =D1, y= NULL)
 
 t.test(x =D2, y= NULL)
 
```
For type 1 the difference between 

**d) Are these changes different between the two types? Give the corresponding 95% confidence interval and p-value**

Since we know there is a more significant increase in the first three months and after a more steady increase or constant migrantion, let us compare the differences in these periods for each type. 





# Q.3 Linear Mixed-effects model

* Consider individuals as random 

```{r}

#Model with only types and time 
model.1 <- lme(nMTPM ~ Month.f*Type.f, random = ~1|ID,
               data = DATA)

# Model with types and time, sex, age, bmi Month.f*Type.f+Sex*BMI+Sex*Month.f
model.2 <- lme(nMTPM ~ Age*Sex+Month.f*Type.f+Sex*BMI+Sex*Month.f, 
               random = ~1|ID,
               data = DATA)

# Exclude Age 
model.3 <- update(model.2, .~.-Age)
summary(model.3)

model.4 <- lme(nMTPM ~ Month.f*Type.f+Sex*BMI+Sex*Month.f, 
               random = ~1|ID,
               data = DATA)
summary(model.4)
anova(model.4)

#remove interaction between Sex and BMI
model.5 <- lme(nMTPM ~ Month.f*Type.f+BMI+Sex*Month.f, 
               random = ~1|ID,
               data = DATA)
summary(model.5)
anova(model.5)

##Check for the use of  "FU.Months""Type" or "Month.f""Type.f"

model.6 <- lme(nMTPM ~ FU.Months*Type+BMI+Sex*FU.Months, 
               random = ~1|ID,
               data = DATA)
summary(model.6)
anova(model.6)

###from the residual plot and qqplot, I think it does not work better than "Month.f""Type.f".

#remove interaction between FU.Months and Type
model.7 <- lme(nMTPM ~ FU.Months+Type+BMI+Sex*FU.Months, 
               random = ~1|ID,
               data = DATA)

model.8 <- lme(nMTPM ~ FU.Months+Type+BMI+Sex*FU.Months, 
                random = ~1|ID,
                weights = varIdent(form = ~ 1 | FU.Months),
                data = DATA)
                
#uncorrelated random intercepts and random slopes model
model.9 <- lme(nMTPM ~ FU.Months+Type+BMI+Sex*FU.Months, 
                random = list(ID = pdDiag(form = ~ FU.Months)),
                weights = varIdent(form = ~ 1 | FU.Months),
                data = DATA)
anova(model.9, model.8)

model..5 <- update(model.5, method = "ML")
model..8 <- update(model.8, method = "ML")
anova(model..5,model..8) #model.5 is better

model..5 <- update(model.5, method = "ML")
model..9 <- update(model.9, method = "ML")
anova(model..5,model..9) #model.5 is better
```

**Between and within subject variation**
```{r}
getVarCov(model..5)

#Between subject variation 
bs <- getVarCov(model..5, type = "random.effects")  [1] #please check because i am not sure about this 

#Within subject variation 
ws <- getVarCov(model..5, type ="conditional")[[1]][1,1] #please check because i am not sure about this  

# Marginal cov matrix 
getVarCov(model..5, type = "marginal", ind = 2)

#Intraclass correlation coefficient of reliability
reliab.coeficcient <- bs/(bs + ws)
reliab.coeficcient
```


##Residual plots 
```{r}
plot(model.5, resid(., type = "p") ~ fitted(.),type = c("p", "smooth"), lwd = 3)
plot(model.7, resid(., type = "p") ~ fitted(.),type = c("p", "smooth"), lwd = 3)

``` 

```{r}
plot(model.8, resid(., type="n") ~ fitted(.),type = c("p", "smooth"), lwd = 3)
plot(model.8, resid(., type="n") ~ fitted(.)|Type.f,type = c("p", "smooth"), lwd = 3)
``` 

**Check for normality**
```{r}
qqnorm(model.8, ~ resid(., type = "p"), abline = c(0, 1))
qqnorm(model.5, ~ resid(., type = "p"), abline = c(0, 1))
# model.5 has better normality based on qqnorm 
``` 


**a) Do both types migrate similarly over time?**



**c) Report the size of the migration change from 3 months to 1 year, 3 months to 2 years, 3 months to 5 years and change from 1 year to 5 year pero operation type and make statistical inference. Given the corresponding 95% confidence interval and p-value.**



**b) Report the differences in the mean migration between the two operation types at each time point**
**Intervals for 95% confidence level** 



**d) Are these changes different between the two types? Give the corresponding 95% confidence interval and p-value**

