---
title: "Prothesis Question 1"
author: "Yuxuan LI, Ariel Liang, Yuying TAN, María Zilli "
date: "May 1st, 2020"
fontsize: 12pt
output:
  html_document:
    df_print: paged
---
```{r , include=TRUE}
load("prostheses.RData")
```

```{r}
DATA
require(pacman)
p_load(lattice, tidyverse, GGally)
```
## Summary of Data
```{r DATA}
summary(DATA)
str(DATA)
```

From the summary we can see that the average age is of patient is of 66.22 years old, with the youngest of 35.6 and the olders is of 80.23; the average Body Mass Index is of 28.32 (lowest 21.20 and max 38.34).There are 160 females and 140 males in the data. 

```{r}
#add Month.f and Type.f as factor 
DATA$Month.f <- factor(DATA$FU.Months, levels = c(0,3,12,24,60))
DATA$Type.f <- factor(DATA$Type, levels = c(1,2))
```

## Plots for Mean Structure and Data exploration 

The following section intends to describe the data, and use graphical techniques to explore the mean structure, the variance structure and the correlation structure. 

**Mean per prothesis type and month** 
```{r}
with(DATA, tapply(nMTPM, list(Month.f, Type), mean, na.rm=TRUE))
```
The means for the Maximum Total Point Motion both increase with both types of prothesis, but the change in mean is larger for type 2. 

**Variance per prothesis type and month** 
```{r}
with(DATA, tapply(nMTPM, list(Month.f, Type), var, na.rm=TRUE))
```


**Plot for individual profiles** 


```{r, echo=FALSE}
xyplot(nMTPM ~ Month.f , groups =ID, data = DATA, xlab = "Month", ylab = "MTPM(mm) ", type = "l")
```

Plot of individual profiles shows an increase in MTPM for almost all individuals from baseline (first day after the cirgury) to 3 months after the operation, after the 3 months we see a fairly continuos trend. 
However, some outliers appear in the graph, particularly individuals which show a great increase in mobility throughout the months and others which show a decerase in mobility. 

**Plot showing individual profiles per type of prothesis** 
```{r, echo=FALSE}
xyplot(nMTPM ~ Month.f | Type.f, groups =ID, data = DATA, xlab = "Month", ylab = "MTPM (mm)", type = "l")
```

For the type 2 we see a steep increase in MTPM 3 months after the cirgury. After the three months more steady however differences in individual profiles.For type 1 there is a slower increase in the profiles and more diverse MTPM. 

**Plot for mean MTPM for each type and period, for both types the MTPM increase**
```{r, echo=FALSE}
xyplot(nMTPM ~ Month.f | Type.f, groups =ID, data = DATA, xlab = "Month", ylab = "MTPM (mm)", type = "l", panel = function(x, y) {
  panel.average(x, y, horizontal = FALSE, col = "blue")
})
```

The mean shows a steapear increase for type 2 than type 1. 


**Age vs. MTPM**
```{r}
p <- ggplot(data = DATA, aes(x = Age, y = nMTPM)) + geom_point( col = "darkblue", alpha = .15)
p+geom_smooth(method = lm )
```



**Body Mas Index vs MTPM** 
```{r}
p1 <- ggplot(data = DATA, aes(x = BMI, y = nMTPM)) + geom_point( col = "darkblue", alpha = .15)
p1 + geom_smooth(method = lm )
```

**Sex vs.MTPM** 
```{r}
plot(nMTPM ~ Sex , data = DATA, type = "l", groups = ID)
```

**Plot per type**
```{r}
m <- ggplot(DATA, aes( x = Month.f, y=nMTPM , group= ID )) 
m + geom_point(aes(colour= Type.f)) 
```

**Plot per Sex and Type**
```{r}
m + geom_point(aes(colour= Type.f)) + facet_grid(.~Sex) 
```

##Conclusions

