---
title: "the answer to Q2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


(a) Do both operation types migrate similarly over time?

A: According to paired t-test, both both operation types do not migrate similarl.

```{r}
DATA$Month.f <- factor(DATA$FU.Months, levels = c(0,3,12,24,60))
DATA$Type.f <- factor(DATA$Type, levels = c(1,2))

t.test(nMTPM~Type.f,data = DATA, paired = TRUE)


```
(b) Report the differences in the mean migration between the two operation types at each time point and make statistical inference.

```{r}
modelq2 <- gls(nMTPM ~ FU.Months*Type.f,
               data = DATA,
               correlation = corCompSymm(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")
summary(modelq2)

# beta0 + beta1 * FU.Months under group type1
# beta0 + beta1 * FU.Months + beta2 + beta3 * FU.Months under group type2
# For those who took type1, the month effect is beta1 * FU.Months
# For those who took type2, the month effect is beta1 * FU.Months + beta3 * FU.Months
# Thus beta3 * FU.Months is the difference in the type at each month

beta3 <- summary(modelq2)$tTable["FU.Months:Type.f2","Value"]
p_value <- summary(modelq2)$tTable["FU.Months:Type.f2","p-value"]
#95% confidence interval
beta3_upp <- beta3 + 1.96 * summary(modelq2)$tTable["FU.Months:Type.f2","Std.Error"] 
beta3_low <- beta3 - 1.96 * summary(modelq2)$tTable["FU.Months:Type.f2","Std.Error"]
beta3inter <- c(beta3_low,beta3_upp)


```

c) Report the size of the migration change from 3 months to 1 year, 3 months to 2 years, 3 months 5 years and change from 1 year to 5 years per operation type and make statistical inference. Give the corresponding 95% confidence interval and p-value.

A:the astimated rate change in type1 group is 0.00798, and 95% intervar is(0.005115998 0.010844256)

```{r}
modelq2 <- gls(nMTPM ~ FU.Months*Type.f,
               data = DATA,
               correlation = corCompSymm(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")
summary(modelq2)

#type 1 change rate
type_1est <- summary(modelq2)$tTable["FU.Months","Value"]
type_1upp <- type_1est +
        1.96*summary(modelq2)$tTable["FU.Months","Std.Error"] 
type_1low <- type_1est -
        1.96*summary(modelq2)$tTable["FU.Months","Std.Error"] 
type_1inter <-c(type_1low, type_1upp)
type_1inter

library(aod)
w.test1 <- wald.test(b = coef(modelq2),
                     Sigma = vcov(modelq2),
                     L = rbind(c(0,1,0,0)))
w.test1

#type 2 change rate 
betas <-coef(modelq2)
V <-vcov(modelq2)
L <-c(0,1,0,1)
type_2est <- c(L %*% betas)
ses <- sqrt(L %*% V %*% L)

type_2upp <-type_2est + 1.96*ses
type_2low <-type_2est - 1.96*ses
type_2inter <- c(type_2low,type_2upp)
type_2inter

w.test2 <- wald.test(b = coef(modelq2),
                     Sigma = vcov(modelq2),
                     L = rbind(c(0,1,0,1)))
w.test2

#type 1 migration change
type_1m31 <- type_1est*(12-3)
type_1m32 <- type_1est*(24-3)
type_1m35 <- type_1est*(60-3)
type_1m15 <- type_1est*(60-12)

#type 2 migration change
type_2m31 <- type_2est*(12-3)
type_2m32 <- type_2est*(24-3)
type_2m35 <- type_2est*(60-3)
type_2m15 <- type_2est*(60-12)

```

(d)Are these changes different between the two types? Give the corresponding 95% confidence interval and p-value

```{r}
Diff_type1 <-c(type_1m31,type_1m32,type_1m35,type_1m15)
Diff_type2 <-c(type_2m31,type_2m32,type_2m35,type_2m15)

t.test(Diff_type1 -Diff_type2)
```


#################################update

data transformation : there are few difference in the residuals plot and qqplot of model17, so we do not consider data transformation 
```{r}
DATA$log_nM <- log(DATA$nMTPM + 1, base = 10)

library(nlme)
model17 <- gls(log_nM ~ Month.f*Type.f+Sex*Month.f-Sex,
               data = DATA,
               correlation = corCompSymm(form = ~ FU.Months | ID),
               weights =  varExp(form = ~ FU.Months),
               na.action = na.exclude, method = "REML")

plot(model17, resid(., type = "p") ~ fitted(.),type = c("p", "smooth"), lwd = 3)
plot(model17, resid(., type="p") ~ fitted(.)|Type.f,type = c("p", "smooth"), lwd = 3)
qqnorm(model17, ~ resid(., type = "p"), abline = c(0, 1))

```

Q3
(a) Do both operation types migrate similarly over time?

```{r}
#the same like Q2(a)
t.test(nMTPM~Type.f,data = DATA, paired = TRUE)

```

(b) Report the differences in the mean migration between the two operation types at each
time point and make statistical inference.

```{r}
modelq3 <- lme(nMTPM ~ FU.Months*Type.f, 
               random = ~1|ID,
               data = DATA)
summary(modelq3)

# beta4 + beta5 * FU.Months under group type1
# beta4 + beta5 * FU.Months + beta6 + beta7 * FU.Months under group type2
# For those who took type1, the month effect is beta5 * FU.Months
# For those who took type2, the month effect is beta5 * FU.Months + beta7 * FU.Months
# Thus beta7 * FU.Months is the difference in the type at each month

beta7 <- summary(modelq3)$tTable["FU.Months:Type.f2","Value"]
p_value <- summary(modelq3)$tTable["FU.Months:Type.f2","p-value"]
#95% confidence interval
beta7_upp <- beta7 + 1.96 * summary(modelq3)$tTable["FU.Months:Type.f2","Std.Error"] 
beta7_low <- beta7 - 1.96 * summary(modelq3)$tTable["FU.Months:Type.f2","Std.Error"]
beta7inter <- c(beta7_low,beta7_upp)
```

(c) Report the size of the migration change from 3 months to 1 year, 3 months to 2 years, 3 months 5 years and change from 1 year to 5 years per operation type and make statistical inference.
```{r}
#type 1 change rate
type_1est <- summary(modelq3)$tTable["FU.Months","Value"]
type_1upp <- type_1est +
        1.96*summary(modelq3)$tTable["FU.Months","Std.Error"] 
type_1low <- type_1est -
        1.96*summary(modelq3)$tTable["FU.Months","Std.Error"] 
type_1inter <-c(type_1low, type_1upp)
type_1inter

library(aod)
w.test3 <- wald.test(b = summary(modelq3)$tTable[,"Value"],
                     Sigma = vcov(modelq3),
                     L = rbind(c(0,1,0,0)))
w.test3

#type 2 change rate 
betas <-summary(modelq3)$tTable[,"Value"]
V <-vcov(modelq3)
L <-c(0,1,0,1)
type_2est <- c(L %*% betas)
ses <- sqrt(L %*% V %*% L)

type_2upp <-type_2est + 1.96*ses
type_2low <-type_2est - 1.96*ses
type_2inter <- c(type_2low,type_2upp)
type_2inter

w.test4 <- wald.test(b = summary(modelq3)$tTable[,"Value"],
                     Sigma = vcov(modelq3),
                     L = rbind(c(0,1,0,1)))
w.test4

#type 1 migration change
type_1m31 <- type_1est*(12-3)
type_1m32 <- type_1est*(24-3)
type_1m35 <- type_1est*(60-3)
type_1m15 <- type_1est*(60-12)

#type 2 migration change
type_2m31 <- type_2est*(12-3)
type_2m32 <- type_2est*(24-3)
type_2m35 <- type_2est*(60-3)
type_2m15 <- type_2est*(60-12)
```

(d)Are these changes different between the two types? Give the corresponding 95% confidence interval and p-value
```{r}
Diff_type1 <-c(type_1m31,type_1m32,type_1m35,type_1m15)
Diff_type2 <-c(type_2m31,type_2m32,type_2m35,type_2m15)

t.test(Diff_type1 -Diff_type2)
```


